What does it take to create with domain-appropriate tools?

### 2020-01-22 convivial-2020/manuscript.tex
DONE

\title{What does it take to create with domain-appropriate tools?}
\subtitle{A case study on the ``OROM'' system}
\author{Joel Jakubovic}
\affiliation{University of Kent, Canterbury}

@@

\newcommand{\joel}[1]{}
\newcommand{\svgel}[1]{\texttt{\textless{}#1\textgreater{}}}
\newcommand{\OSFA}{One-Size-Fits-All}

\begin{abstract}
There is a One-Size-Fits-Allquality to languages, APIs and even programming itself. Whether you're making a mobile game or a scientific simulation, you will be using a text-based language with similar devices for structuring your code. This is a source of artificial difficulty in creating, understanding, and modifying software systems. No matter the domain, the author's design needs encoding into a form that does not resemble it.

@@

One would have thought that software-creation acts like hypothetical
super-intelligent AI. That is: even though we start from a primitive
base in the 50s (or even today), there would surely be a recursive
process of self-improvement, building better software-creation tools
with the existing ones, until an ``expressivity singularity'' where
software becomes a workable material as described.

@@

Surely the trick is to \emph{use} coding to build something \emph{better
than it}. And then use that, to build something even better. But there
is an enormous breadth and depth of philosophies here, along with many
failed historical attempts to do better --- or at least, ones that
failed to catch on. But even worse than this is that in my very
\emph{language} here I am making the same mistake as the text editor ---
speaking in unqualified terms of ``better'' and ``worse'' as if there
really is a One-Size-Fits-Allsolution to software creation!

Of course what we \emph{really} want is the ability for people to create
\emph{in the way that they think is best} in their particular context
--- to equip them to feasibly create the tools that suit them for the
thing they want to make. And second-order tools that suit them for
making the first-order tools, etc. It would do no good to replace
text-imperialism with anything-else-imperialism, which is one
interpretation of calls for alternatives.

This dream goes beyond the familiar sense of what constitutes a
``craft'', as far as a strong melding of tool and material. Parallels
can be drawn with industrialisation and a strong division of labour: the
community as a whole produces its higher-order tools, but currently no
single person can have the same autonomy. A (future) software craft
could be expected to give this power to \emph{individuals}, instead of
the community alone. Whenever there are many small specialities
(e.g.~languages, tools, or subject areas) each serving many clients, the
One-Size-Fits-Allstyle is the best one can hope for. Adaptation to
individual preferences and idiosyncrasies is only feasible when those
individuals can do it themselves.

@@

When I first read the paper Open, Reusable Object Models, I was hooked
on its idea of a small but expressive starting system that could be
self-improved into anything. It describes a late-bound\footnote{Fewer
  commitments; more things determined by runtime conditions},
Smalltalk-style\footnote{OOP with more emphasis on object instances and
  messaging (methods) as in a distributed system; less emphasis on class
  hierarchies implementing traditional data structures} objects and
messaging environment that the authors call ``Id'' --- but which I refer
to as OROM, for pronunciation and Googleability.

@@

The paper itself consists of mostly prose, several code listings, and
full C sources for a sample implementation at the end. It also provides
several diagrams. But to understand it, I repeatedly found myself
drawing \emph{extra} diagrams. For example, the first acts of the
running system boil down to initialising the three or so objects. This
consists of allocating memory, interpreting it as a C struct and then
filling in fields in a mundane manner. I had great difficulty following
the specifics in my head, but when I drew tables in the style of their
diagrams I readily saw what what going on. I personally would have
preferred these to have been in the paper in the first place, but I am
not necessarily representative of those who read it. The
One-Size-Fits-Allapproach is perhaps unavoidable for static media like
print.

@@

Thus was my natural representation decided. My first attempt to make it
a reality was a partial success: a webpage made of HTML tables, evolved
via JavaScript.

@@

To emphasise the tendency of OROM objects to be visualised as key-value
mappings, I will refer to them as \emph{obj-dicts}. Figure
\ref{fig:orom-html} shows the OROM/HTML implementation, in which they
take the form of HTML tables.

@@

For each obj-dict I used a two-column
\texttt{\textless{}table\textgreater{}} within a
\texttt{\textless{}div\textgreater{}}, matching the diagrams, and gave
the rows CSS class names matching the keys for easy lookup. Sending
messages relied on the JavaScript console, but existing values in text
boxes (such as method implementations) could be edited directly. Also
thanks to the browser, I could easily make text boxes and
\texttt{\textless{}div\textgreater{}}s resizable as necessary.

However, the choice of ordinary HTML as a substrate proved rather
two-edged. The browser requires many features to do its job of rendering
complex web pages. And sadly, as its client, I could only make use of
those capabilities which the W3C had decided, at the time of authorship,
were worth the effort exposing in JavaScript. For anything else, the
browser is a black box, and this was very frustrating in the following
case.

@@

A key aspect of the OROM system is that there is an object \emph{graph}.
That is, obj-dicts can have entries pointing to other obj-dicts, without
restriction to a tree structure. Drawing arrows between things to denote
this is a no-brainer, and I wanted it in my substrate for OROM to match
the diagrams.

A small problem to surmount first: even though I could hijack the
\texttt{\textless{}table\textgreater{}} for its display properties, what
element could I hijack to make arrows between arbitrary points? Luckily,
there was SVG at my disposal, which could be persuaded to display over
the HTML. However, another key feature of my intended substrate was to
be able to rearrange and resize the boxes as desired. So I would also
need to detect changes to the position and size of an element.

@@

\hypertarget{practicality-vs.-high-aesthetics}{%
\subsection{Practicality vs.~High
Aesthetics}\label{practicality-vs.-high-aesthetics}}

There is the view that insists on provably-correct code written with
strange Unicode symbols, and at the opposite end there is the attitude
of ``get the job done'', high style be damned. Each has its merits, but
for my task I quickly saw the latter route to be most suitable.

My favourite example of this has been the ``encoding'' of obj-dicts as
DOM trees containing particular arrangements of children. In OROM/HTML,
I would gleefully get/set state by calling code like this:

\begin{lstlisting}[language=JavaScript]
Entity.prototype.addState = function(key) {
  let tr = document.createElement('tr');
  tr.setAttribute('class', key);

  let td = document.createElement('td');
  td.setAttribute('class', 'key');
  td.textContent = key;
  tr.appendChild(td);
  ...
\end{lstlisting}

and in OROM/SVG the story is similar. It takes advantage of the powerful
developer tools included in browsers (e.g.~the DOM inspector), and
avoids the need to manually synchronise separate Model and View for such
encoded state. With a little work, it also allowed the system to be
saved and restored --- see Section \ref{persistence}. I consider this as
practical, expedient, ``hijacking'' of these substrates because it
either goes against their intended use (HTML tables for presentation
rather than semantics) or are simply ``odd'' (using certain arrangements
of shapes to ``encode'' machine-readable semantics).

Contrast this with an approach that had to re-synthesise bits of the DOM
whenever something changed. Admittedly this does still happen to a
certain extent, but at least it was minimised where obvious.

Another example of practicality was in the issue of positioning and
sizing. In Section \ref{nut-cracking-with-sledgehammers} I mentioned
that I was able to sidestep some layout work by offloading it to the
user. While building the system I tended to want to draw boxes into
existence, and for that my brain's existing aesthetic algorithms can
decide where they go. There are areas where the machine must be able to
figure this out itself --- I don't want to be constantly bothered by
prompts to place boxes as they are allocated by some running code ---
but for one-offs it is a very convenient way to do a little less
yak-shearing.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{../orom-svg.png}
  \caption{OROM/SVG: Obj-dicts are moveable and resizable nested boxes,
           referencing each other via real arrows.\label{fig:orom-svg}}
\end{figure}

In this version, shown in Figure \ref{fig:orom-svg}, obj-dicts are
encoded as nested SVG \texttt{\textless{}rect\textgreater{}}s and other
elements, reminiscent of diSessa's Boxer. This was a significant
departure from the table representation, and even though SVG supports
(some) nested HTML via \texttt{\textless{}foreignObject\textgreater{}},
I actually preferred the possibility of multiple levels of nesting.

OROM/SVG more or less realises my desired substrate for implementing
OROM. Needless to say, this version was far more challenging and took
much longer to reach a satisfactory state. However, it is precisely this
drudgery that brings me to a better understanding of this paper's
question: \emph{what has it taken?} I shall discuss this in the form of
broad patterns or themes that stand out to me from my development
experience, hammered home by these OROM projects.

@@

The ``common case'' of software present in our daily lives shares
certain properties, such as being graphical and interactive. Such are
expectations that ``end-users'' hold, consigned as they are to merely
\emph{consume} what programmers give them. But if we want to make
``programming'' more like ``using'', such expectations on \emph{use} at
least need to be acknowledged. In this section, I will present these and
other seemingly \emph{inevitable} demands of normal software,
exemplified by OROM/SVG. I will comment how our programming platforms
measure up to the task of helping us meet them, including my chosen
platform of JavaScript and Web technologies.

@@

On this requirement, SVG fits the bill very well. Although it is not
part of the JavaScript language \emph{per se}, it is a standard and
widely supported technology of the Web \emph{platform}. We can observe
that anyone with a browser \emph{in principle} has access to a powerful
vector graphics editor --- just one with no GUI. I will return to this
at the end (Section ??).

The SVG tree keeps the nice properties of the DOM, such as updating the
display when shape parameters are changed. This is well-adapted to
``I/O-bound'' software like mine, where things change only in response
to user input. If I wanted animation, this boils down to a regular
``advance simulation'' signal, and would require setting up some
rendering loop. Alternatively, there is the W3C's chosen ontology of CSS
animations, but see Section \ref{context-appropriate-ontologies}.

@@

All humans learn a basic set of expectations about the things they see
around them. Some of these, such as ``things fall down'', are not
generally appropriate to software UIs --- perhaps because the screen has
a role in our lives a more of a table work surface rather than a
vertical wall, even if it is vertical in real life.

The level of physics in software tends to not involve force, or mass, or
very much at all, merely position and space; we could call it
``geometric physics''. One thing that all usable software must do, for
example, is not crush many visually complex shapes, such as lines of
text, into the same region, since it becomes unreadable. Such concepts
of ``solid objects do not intersect'' or ``only things at different
layers may overlap'' are basic rules inherited from the real world of
graphical presentation.

@@

\begin{itemize}
\tightlist
\item
  This number range 0000-FFFF is like a space, numbers = points
\item
  Every point has at most one owner block
\item
  These blocks are contiguous, finite ranges (i.e.~1D boxes)
\end{itemize}

Hence the boilerplate involved to realise this in any domain with
something resembling space. Far from being a niche topic in games and
graphics, spatial partitioning algorithms and data structures have
surprising relevance to more ordinary software. Both memory allocation
and graphical layout are essential to today's; shame that only one of
those has been recognised as such and entered the runtime of modern
languages.

@@

When you have both a screen and a pointing device (e.g.~touch or mouse),
immediately it becomes worth having ways to move things around in at
least a minimally realistic way. We can debate the appropriateness of
Direct Manipulation (DM) for various situations. But it does make a lot
of sense in simple cases, such moving around subdivisions of space
(e.g.~windows) or elements of a graphical design.

In OROM, the obvious candidate for this is the obj-dicts, plus all
nested boxes in OROM/SVG. If I move the top-level rect, then I expect
its children to move with it. This is simply the translational physics
of rigid bodies: this set of points all move together. Of course, proper
rigid bodies might also rotate and have mass, but this is usually
undesirable for UI elements.

Translational rigidity can be expressed as the points X and Y always
having the same displacement from each other. Or, when one point is
moved, the rest also move by the same delta. This is a problem of
preserving the relationship over time, which was a significant area of
OROM/SVG.

@@

What more could there be? Well, in \emph{every} software system there
are certain rules, or ``invariants'' of internal consistency, such as
``translational rigidity'' above. Often, changes to any part of the
system are permissible, but only if connected or dependent parts of the
state change in response.

The job of keeping track of who depends on whom can fall either on the
programmer, or the computer. If the programmer has to do this, they can
only go so far managing and simulating in their head. As systems grow
more complex, it is only natural to try and make the computer more
intelligent to do this work. What I am building up to is that whenever
we (or I) consider constraints, or ``reactive'' programming, or the
``Observer'' pattern as things you only wheel out on special occasions,
we only deceive ourselves into doing the same work less explicitly. It
seems that such ``live state'' should be the expected common case for
software development.

@@

Getting an object to follow the mouse pointer, for example when
dragging, is conceptually very simple: an ``always equal'' relation. In
OROM/SVG founded on live-state, this can be expressed in much the same
way:

@@

(By default, an Observable A responds to a change from Observable B by
adopting B's new value.)

@@

Speaking about vector graphics, physics, layout and constraint
maintenance might give the impression of high \emph{conceptual}
complexity at the heart of even simple software. This is not quite true,
which makes it all the worse that there is yet still immense
\emph{incidental} or implementation complexity.

We are conditioned to only think of these in their most general forms.
The vector graphics I use in OROM/SVG are just rects, lines, circles and
text; a fraction of the full capability of SVG. The ``geometric
physics'' I use is dwarfed by fully general 2D or 3D physics engines.
The only layout algorithm I had the patience to implement was a simple
way to expand a list of boxes to fit in a new child at the bottom; the
affordance to place and size boxes \emph{manually} staves off the rest
for the time being! Yet search for material on layout algorithms and it
can seem like Fully General Linear Inequality Solvers like Cassowary are
all there is.

The inevitable requirements I suggest here, do \emph{not} necessitate
Fully General anything. In fact, such representations would make it
\emph{more} cumbersome to express what I wanted in OROM. As the old
wisdom goes, there's no point expressing a simple regex search as an
arbitrary Turing machine; my boxes don't \emph{have} a moment of inertia
or a mass and I don't \emph{need} a linear optimisation solver for my
space management --- for the time being. Under the theme of
domain-appropriate tools, it is worth designing interfaces for
smaller-scale instances of these areas and exploring their capabilities.

@@

The message of the previous section is that existing platforms are often
at the wrong level of abstraction for the requirements of common
software. I recognise that they are reasonably well-adapted to batch
mode file I/O tasks, but that they fail for the \emph{common case} is a
problem. It necessitates largely the same ``boilerplate'' setup per
project just to get basic functionality:

@@

It is true, we already have a term for bringing into existence some
software feature that isn't already there: programming. The difference
is that polyfilling is about filling in \emph{boilerplate},
i.e.~functionality \emph{that should have been there in the first place,
but isn't}. This injects some subjectivity and value judgement into the
term. But ultimately, all platforms are designed with certain features
``out of the box'' and leave other features to be implemented ``if
needed''. And when a problem domain ``requires batteries'' yet the tools
available say ``batteries not included'', it is quite reasonable to want
the batteries to be part of the platform.

This section covers the boilerplate I found myself polyfilling in
OROM/SVG.

Is to bring things into the mindset / ontology of the specific software.
For example, wrapping the official Web conventions for event handling in
the Observable infrastructure, or wrapping DOM nodes with ``user data''
objects. The latter case, takes advantage of JavaScript's ability to
simply ``add on'' new fields to objects, even objects of Web APIs, to
``annotate'' the DOM node with a JavaScript helper. Very advantageous to
permit this in JavaScript. In another language like C, such would
require saving the address in a lookup table elsewhere.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.75\textwidth]{../rods.png}
  \caption{The yellow rods on the box border are ``half-rigid'': they transmit horizontal or vertical changes, but not both. The green rod in the top-left absorbs all changes from an endpoint into itself, visualising the displacement vector of the child box from its parent. The remaining red rods are fully rigid. A line of CSS can show this ``scaffolding'' which is normally hidden for users. \label{fig:rods}}
\end{figure}

@@

I did manage to surmount this through kludging and compromise. But I do
not know whether these problems are merely a consequence of some design
decision I could change to escape it, or if they are intrinsic to my
(modest) UI goal.

@@

All child elements of a node transform with it, so already SVG has baked
in a basic facility for translational rigidity. This is only available
as a tree hierarchy, but it is still useful. However, it conflicts with
my early decision to have Point objects all share the global co-ordinate
system (this was to ensure that simple relations, such as a point
following the mouse pointer, are not infuriating to express.) Still, it
was necessary in the case of certain elements --- especially those which
must transcend the tree structure altogether, like arrows between boxes
--- to bite this bullet, one way or another.

\begin{quote}
Maybe mention that this is caused by the mismatch between the substrate
and the programming system (substrate is tree, but programming is
graphs)?
\end{quote}

Again, I return to how we tend to work things out in the freedom of
paper. Co-ordinate systems, here merely positionally displaced, have
their origins here and there and have vectors between them. The rods
thus far let me visually express relations between global Points; now
was a question of expressing one global Point as a displacement from
another (the \texttt{\textless{}g\textgreater{}} transform). New rod
Observables \texttt{p2\_from\_p1} and \texttt{p1\_from\_p2} do the
vector subtraction, which can then be propagated as local co-ordinates
to children. It is nicer to express the relation (as well as see it!)
this way (Figure \ref{fig:rods}).

@@

Each API has its own conventions, including a way of naming and
structuring expressions --- an ontology. The One-Size-Fits-Allapproach
is exemplified in such interfaces. For example, in SVG we express a
rectangle by
\texttt{\textless{}rect x="10" y="10" width="600" height="400"\textgreater{}}.
The SVG specification \emph{defines} to the user that a rect \emph{is} a
top-left corner, a width, and a height --- and that's it. However, this
SVG-approved parametrisation of a rectangle is far from the only
one\href{https://www.shift-society.org/salon/papers/2018/critiques/critique-semprola.pdf}{1},
and thus is, unsurprisingly, ill-fitted to some contexts.

@@

Another example is to be found in the DOM's event listener model.
Conceptually, many ``events'' are in fact changes to the state of some
physical (or conceptual) device. And both keyboard keys and mouse
buttons, for example, have two states --- pressed, not pressed --- which
ought to make them \emph{interchangeable} to this extent. This is the
logic behind ``key mapping''\footnote{For example, the player can
  re-assign the action ``shoot'' from its default mouse button to a
  keyboard key, or another action from keyboard to mouse, as it suits
  them.} in PC games. But the official ontology of the Web makes it
nontrivial to do this.

>>>>

\title{What does it take to create with domain-appropriate tools?}
\subtitle{A case study on the ``OROM'' system}
\author[ ]{Joel Jakubovic}
\affiliation[ ]{University of Kent, Canterbury}

@@

\newcommand{\joel}[1]{}
\newcommand{\svgel}[1]{\texttt{\textless{}#1\textgreater{}}}
\newcommand{\OSFA}{One-Size-Fits-All }

\begin{abstract}
There is a One-Size-Fits-All quality to languages, APIs and even programming itself. Whether you're making a mobile game or a scientific simulation, you will be using a text-based language with similar devices for structuring your code. This is a source of artificial difficulty in creating, understanding, and modifying software systems. No matter the domain, the author's design needs encoding into a form that does not resemble it.

@@

One would have thought that software-creation acts like hypothetical
super-intelligent Artificial Intelligence (AI). That is: even though we
start from a primitive base in the 50s (or even today), there would
surely be a recursive process of self-improvement, building better
software-creation tools with the existing ones, until an ``expressivity
singularity'' where software becomes a workable material as described.

@@

Surely the trick is to \emph{use} coding to build something \emph{better
than it}. And then use that, to build something even better. But there
is an enormous breadth and depth of philosophies here, along with many
failed historical attempts to do better --- or at least, ones that
failed to catch on. And even worse than this is that in my very
\emph{language} here I am making the same mistake as the text editor ---
speaking in unqualified terms of ``better'' and ``worse'' as if there
really is a One-Size-Fits-All solution to software creation!

Of course what we \emph{really} want is the ability for people to create
\emph{in the way that they think is best} in their particular context
--- to equip them to feasibly create the tools that suit them for the
thing they want to make. And second-order tools that suit them for
making the first-order tools, and so on. It would do no good to replace
text-imperialism with anything-else-imperialism, which is one
interpretation of calls for alternatives.

This dream goes beyond the familiar sense of what constitutes a
``craft'', as far as a strong melding of tool and material. Parallels
can be drawn with industrialisation and a strong division of labour: the
community as a whole produces its higher-order tools, but currently no
single person can have the same autonomy. A (future) software craft
could be expected to give this power to \emph{individuals}, instead of
the community alone. Whenever there are many small specialities
(e.g.~languages, tools, or subject areas) each serving many clients, the
One-Size-Fits-All style is the best one can hope for. Adaptation to
individual preferences and idiosyncrasies is only feasible when those
individuals can do it themselves.

@@

When I first read the paper ``Open, Reusable Object Models''
\cite{OROM}, I was hooked on its idea of a small but expressive starting
system that could be self-improved into anything. It describes a
late-bound\footnote{Fewer commitments; more things determined by runtime
  conditions}, Smalltalk-style\footnote{OOP with more emphasis on object
  instances and messaging (methods) as in a distributed system; less
  emphasis on class hierarchies implementing traditional data structures}
objects and messaging environment that the authors call ``Id'' --- but I
shall refer to it as OROM, for pronunciation and Googleability.

@@

The paper itself consists of mostly prose, several code listings, and
full C sources for a sample implementation at the end. It also provides
several diagrams. But to understand it, I repeatedly found myself
drawing \emph{extra} diagrams. For example, the first acts of the
running system boil down to initialising the three or so objects. This
consists of allocating memory, interpreting it as a C struct and then
filling in fields in a mundane manner. I had great difficulty following
the specifics in my head, but when I drew tables in the style of their
diagrams I readily saw what what going on. I personally would have
preferred these to have been in the paper in the first place, but I am
not necessarily representative of those who read it. The
One-Size-Fits-All approach is perhaps unavoidable for static media like
print.

@@

Thus was my natural representation decided. My first attempt to make it
a reality was a partial success: a webpage made of HTML tables, evolved
via JavaScript (JS).

@@

To emphasise the tendency of OROM objects to be visualised as key-value
mappings, I will refer to them as \emph{obj-dicts}. Figure
\ref{fig:orom-html} shows the OROM/HTML implementation \cite{orom-html},
in which they take the form of HTML tables.

@@

I used a two-column \texttt{\textless{}table\textgreater{}} within a
\texttt{\textless{}div\textgreater{}} for each obj-dict (matching the
diagrams) and gave the rows CSS class names matching the keys (for easy
lookup). Sending messages relied on the JS console, but existing values
in text boxes (such as method implementations) could be edited directly.

This choice of ordinary HTML as a substrate, however, proved rather
two-edged. The browser requires many features to do its job of rendering
complex web pages. And sadly, as its client, I could only make use of
those capabilities which the W3C had decided, at the time of authorship,
were worth the effort exposing in JavaScript. For anything else, the
browser is a black box, and this was very frustrating in the following
case.

@@

A key aspect of the OROM system is that there is an object \emph{graph}.
That is, obj-dicts can have entries pointing to other obj-dicts, without
restriction to a tree structure. Drawing arrows to denote this is a
no-brainer (and easy on paper), so I wanted it in my substrate for OROM.

A small problem to surmount first: even though I could hijack the
\texttt{\textless{}table\textgreater{}} for its display properties, what
element could I hijack to make arrows between arbitrary points? Luckily,
there were Scalable Vector Graphics (SVG) at my disposal, which could be
persuaded to display \texttt{\textless{}line\textgreater{}}s over the
\texttt{\textless{}div\textgreater{}}s. But another key feature of my
intended substrate was to be able to rearrange and resize the boxes. So
I would also need to detect changes to the position and size of an
element.

@@

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{../orom-svg.png}
  \caption{OROM/SVG: Obj-dicts are moveable and resizable nested boxes,
           referencing each other via \emph{real arrows}.\label{fig:orom-svg}}
\end{figure}

In OROM/SVG, shown in Figure \ref{fig:orom-svg}, obj-dicts are encoded
as nested SVG \texttt{\textless{}rect\textgreater{}}s and other
elements, reminiscent of diSessa's Boxer\cite{boxer}. This was a
significant departure from the table representation, and even though SVG
supports (some) nested HTML via
\texttt{\textless{}foreignObject\textgreater{}}, I actually preferred
the possibility of multiple levels of nesting.

Still, this version was far more challenging and took much longer to
reach a satisfactory state. Yet it is precisely this drudgery that
brings me to a better understanding of this paper's question: \emph{what
has it taken?} I shall discuss this in the form of broad patterns or
themes that stand out to me from my development experience, hammered
home by these OROM projects.

@@

The ``common case'' of software present in our daily lives shares
certain properties, such as being graphical and interactive. Such are
expectations that ``end-users'' hold, consigned as they are to merely
\emph{consume} what programmers give them. But if we want to make
``programming'' more like ``using'', such expectations on \emph{use} at
least need to be acknowledged.

In this section, I will present these and other seemingly
\emph{inevitable} demands of normal software, exemplified by OROM/SVG. I
will comment on how our programming platforms measure up to the task,
including my chosen platform of JavaScript and Web technologies.

@@

On this requirement, SVG fits the bill very well. Although it is not
part of the JavaScript language \emph{per se}, it is a standard and
widely supported technology of the Web \emph{platform}. We can observe
that anyone with a browser \emph{in principle} has access to a powerful
vector graphics editor --- just one with no GUI.

The SVG tree has the nice properties of the DOM, such as updating the
display when shape parameters are changed. This is well-adapted to
``I/O-bound'' software like mine, where things change only in response
to user input. If I wanted animation, this would boil down to a regular
``advance simulation'' signal, and would require setting up some
rendering loop. Alternatively, there is the W3C's chosen ontology of CSS
animations, but see Section \ref{context-appropriate-ontologies}.

@@

All humans learn a basic set of expectations about the things they see
around them. Some of these, such as ``things fall down'', are not
generally appropriate to software UIs --- perhaps because the screen has
a role in our lives a more of a table work surface rather than a
vertical wall, even if it is vertical in real life. The level of physics
in software tends to not involve force, or mass, or very much at all,
merely position and space; we could call it ``geometric physics''.

One thing that all usable software must do, for example, is avoid
crushing many visually complex shapes, such as lines of text, into the
same (unreadable) region. Such concepts of ``solid objects do not
intersect'' or ``only things at different layers may overlap'' are basic
rules inherited from the real world of graphical presentation.

@@

\begin{itemize}
\tightlist
\item
  This number range \texttt{0000}--\texttt{FFFF} is like a space,
  addresses = points
\item
  Every point has at most one owner block
\item
  These blocks are contiguous, finite ranges (i.e.~1D boxes)
\end{itemize}

Hence the boilerplate involved to realise this in any domain with
something resembling space. Far from being a niche topic in games and
graphics, spatial partitioning algorithms and data structures have
surprising relevance to more ordinary software. Both memory allocation
and graphical layout are essential to today's; shame that only one of
those has been recognised as such --- and entered the runtime of modern
languages.

@@

When you have both a screen and a pointing device (e.g.~touch or mouse),
immediately it becomes worth having ways to move things around in at
least a minimally realistic way. We can debate the appropriateness of
Direct Manipulation for various situations. But it does make a lot of
sense in simple cases, such moving around subdivisions of space
(e.g.~windows) or elements of a graphical design.

In OROM, the obvious candidate for this is the obj-dicts, plus all
nested boxes in OROM/SVG. If I move the top-level rect, then I expect
its children to move with it. This is simply the translational physics
of \textbf{rigid bodies}: ``this set of points all move together''. Of
course, proper rigid bodies might also rotate and have mass, but this is
usually undesirable for UI elements.

Translational rigidity can be expressed as the points X and Y always
having the same displacement from each other. Or, when one point is
moved, the rest also move by the same delta. This is a problem of
\textbf{preserving the relationship over time}, which was a significant
area of OROM/SVG.

@@

What more could there be? Well, in \emph{every} software system there
are certain rules, or ``invariants'' of \textbf{internal consistency},
such as ``translational rigidity'' above. Often, changes to any part of
the system are permissible, but only if connected or dependent parts of
the state change in response.

The job of keeping track of who depends on whom can fall either on the
programmer, or the computer. If the programmer has to do this, they can
only go so far managing and simulating in their head. As systems grow
more complex, it is only natural to try and make the computer more
intelligent to do this work. What I am building up to is that whenever
we (or I) consider constraints, ``reactive'' programming, or the
``Observer'' pattern as things you only wheel out \emph{on special
occasions}, we only deceive ourselves into doing the \emph{same work}
less explicitly. It seems that such ``live state'' should be the
expected common case for software development.

@@

Getting an object to follow the mouse pointer (e.g.~when dragging) is
\emph{conceptually} very simple: an ``always equal'' relation. In
OROM/SVG founded on live-state, this can be expressed in much the same
way:

@@

By default, an Observable A responds to a change from Observable B by
adopting B's new value. So by subscribing the object's position to the
pointer's, pointer movements copy the new position to the object.

@@

Speaking about vector graphics, physics, layout and constraint
maintenance might give the impression of high \emph{conceptual}
complexity at the heart of even simple software. This is not quite true,
which makes it all the worse that there is yet still immense
\emph{implementation} complexity.

We are conditioned to only think of these in their most general forms.
But the ``vector graphics'' I use in OROM/SVG are just rects, lines,
circles and text; a fraction of the full capability of SVG. The
``geometric physics'' I use is dwarfed by fully general 2D or 3D physics
engines. The only layout algorithm I had the patience to implement was a
simple way to expand a list of boxes to fit in a new child at the
bottom. The affordance to place and size boxes \emph{manually} is a
convenient substitute, when required infrequently. Yet search for
material on layout algorithms, and it can seem like Fully General Linear
Inequality Solvers like Cassowary \cite{cassowary} are all there is.

The inevitable requirements I suggest here, do \emph{not} necessitate
Fully General anything. In fact, such generality might make it
\emph{more} cumbersome to express what I wanted in OROM. As the old
wisdom goes, there's no point expressing a simple regex search as an
arbitrary Turing machine; my boxes don't \emph{have} a moment of inertia
or a mass and I don't \emph{need} a linear optimisation solver for my
space management --- for the time being. Under the theme of
domain-appropriate tools, I think it is worth designing interfaces for
\emph{smaller-scale} instances of these areas and exploring what they
can express.

@@

The message of the previous section is that existing platforms are often
at the wrong level of abstraction for the requirements of common
software. I recognise that they are reasonably well-adapted to batch
mode file I/O tasks, but that they fail for the \emph{common case} is a
problem. There is largely the same setup per project just to get basic
functionality:

@@

It is true, we already have a term for bringing into existence some
software feature that isn't already there: programming. The difference
is that polyfilling is about filling in \emph{boilerplate},
i.e.~functionality \emph{that should have been there in the first place,
but isn't}. This injects some subjectivity and value judgement into the
term. But ultimately, all platforms are designed with certain features
``out of the box'' and leave other features for us to implement ``if
needed''.

When the \emph{normal} problem domain ``requires batteries'', yet the
tools available say ``batteries not included'', it is quite reasonable
to want the batteries to be part of the platform. This section details
the ``batteries not included'' I found myself polyfilling in OROM/SVG.

\hypertarget{wrapping-and-user-data}{%
\subsection{Wrapping and ``user data''}\label{wrapping-and-user-data}}

The classic example of polyfilling is finding that JavaScript arrays do
not support some familiar operation. For example, this used to be the
case with the \texttt{map} function. Yet JavaScript lets you
\emph{directly augment} the \texttt{Array} class with the new operation,
even though \texttt{Array} is ``part of the language'':

\begin{lstlisting}
Array.prototype.map = function() { ... }
\end{lstlisting}

One can do this to all objects, even those of Web APIs. For example, I
use it to connect SVG nodes to the OROM system: when a
\texttt{\textless{}rect\textgreater{}} is created, it is ``annotated''
with a \texttt{userData} pointer to the box that it is part of. Then,
when clicked, the box can be found immediately.

This could be described as the box ``wrapping'' the
\texttt{\textless{}rect\textgreater{}}, but risks impliying that the
\texttt{\textless{}rect\textgreater{}} merely needs to \emph{live
inside}, or \emph{be referenced by}, the box. In this case, the
\emph{reverse} association is also needed, because clicks can only enter
the system \emph{through} SVG in the first place. What is significant
here is that JavaScript lets us add such a link to an entity that
\emph{we did not design}.

This term, ``user data'', often crops up in libraries with the same
idea: to associate arbitrary data with an API object. But in languages
that do not permit such ``annotation'', this has to be explicitly
\emph{designed in} ahead of time. If it wasn't, then it has to be done
in a roundabout way --- for example, a separate \texttt{userData} lookup
table, indexed by the memory address of the API object.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{../rods.png}
  \caption{The yellow rods on the box border are ``half-rigid'': they transmit horizontal or vertical changes, but not both. The green rod in the top-left absorbs all changes from an endpoint into itself, visualising the displacement vector of the child box from its parent. The remaining red rods are fully rigid. A line of CSS can show this ``scaffolding'' which is normally hidden for users. \label{fig:rods}}
\end{figure}

@@

I had to shelve this investigation in the interest of continuing with
the rest of the system, but I did manage to surmount this through
kludging and compromise. So I do not know whether these problems are
merely a consequence of some design decision I could change to escape
it, or if they are intrinsic to my (modest) UI goal.

@@

All child elements of a node transform with it, so already SVG has baked
in a basic facility for translational rigidity. This is only available
as a tree hierarchy\footnote{This highlights the mismatch between the
  tree-based DOM and any system that is graph-structured.}, but it is
still useful. However, it conflicts with my early decision to have Point
objects all share the global co-ordinate system (this was to ensure that
simple relations, such as a point following the mouse pointer, are not
infuriating to express.) Still, it was necessary in the case of certain
elements --- especially those which must transcend the tree structure
altogether, like arrows between boxes --- to bite this bullet, one way
or another.

Again, I return to how we tend to work things out in the freedom of
paper. Co-ordinate systems, here merely positionally displaced, have
their origins here and there and have vectors between them. The rods
thus far let me visually express relations between global Points; now
was a question of expressing one global Point as a displacement from
another (the \texttt{\textless{}g\textgreater{}} transform). New rod
Observables \texttt{p2\_from\_p1} and \texttt{p1\_from\_p2} do the
vector subtraction, which can then be propagated as local co-ordinates
to children. It is nice to express the relation (as well as see it!)
this way (Figure \ref{fig:rods}).

@@

Each API has its own conventions, including a way of naming and
structuring expressions --- an ontology \cite{crit-semprola}. The
One-Size-Fits-All approach is exemplified in such interfaces. For
example, in SVG we express a rectangle by
\texttt{\textless{}rect x="10" y="10" width="600" height="400"\textgreater{}}.
The SVG specification \emph{defines} to the user that a rect simply
\textbf{is} a top-left corner, a width, and a height --- and that's it.
However, this SVG-approved parametrisation of a rectangle is far from
the only one, and thus is, unsurprisingly, ill-fitted to some contexts.

@@

Another example is to be found in the DOM's event listener model.
Conceptually, many ``events'' are in fact changes to the state of some
physical device. And both keyboard keys and mouse buttons, for example,
have two states --- pressed, not pressed --- which ought to make them
\emph{interchangeable} to some extent. Indeed, this is why there is such
a thing as ``key mapping''\footnote{For example, the player can
  re-assign the action ``shoot'' from its default mouse button to a
  keyboard key, or another action from keyboard to mouse, as it suits
  them.} in PC games. But the official ontology of the Web makes it
nontrivial to do this.

To start with, the situation is modelled not as the \emph{changing} of
some time-varying property, but instead as a sort of Cartesian product
of event listeners. Rather than, say, a piece of live-state for the left
mouse button (LMB), we get \texttt{onmousedown} and \texttt{onmouseup}.
Thankfully we do not have \texttt{onAdown}/\texttt{onAup},
\texttt{onBdown/onBup}, \ldots{} all the way to
\texttt{onZdown}/\texttt{onZup}, but only
\texttt{onkeydown}/\texttt{onkeyup}. Yet this is just one of the many
possible ways to slice this 3-dimensional\footnote{Device (mouse,
  keyboard), sub-device (button, key), state (up, down)} space.
